Causal Bandits: Learning Good Interventions via Causal Inference

### 2016-02-05 causal-bandit/icml-2015/experiments.tex
final version of figs for experiments
fixes

\subsection{Parallel bandits problem}

For the parallel bandit problem, we use a model in which, unknown to the algorithms, $Y$ depends only on a single variable.
\eq{
Y_t \sim \begin{cases}
\bernoulli(\frac{1}{2}+\epsilon) & \text{if } X_1 = 1 \\
\bernoulli(\frac{1}{2}-\frac{q_1}{1-q_1}\epsilon) & \text{otherwise}\,.
\end{cases}
}

@@

\begin{figure}[h]
\centering
\includegraphics[width=.5\textwidth]{exp_regret_vs_m_N50_T400_s1000_20160205_0910.pdf}
\caption{Simple regret vs $m(\boldsymbol{q})$ for fixed horizon $T=400$ and number of variables $N = 50$}
\label{fig:simple_vs_m}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=.5\textwidth]{exp_regret_vs_T_N50_a4_s10000_20160205_1532}
\caption{Simple regret vs horizon, $T$, with $N = 50$, $m=2$ and fixed $\epsilon = .3$}
\label{fig:simple_vs_T}
\end{figure}

>>>>

We compare Algorithms \ref{alg:simple} and \ref{alg:general} with Successive Elimination on the parallel bandit problem
under a variety of conditions, including where the importance weighted estimator used by Algorithm \ref{alg:general} is not truncated,
which is justified in this setting by Remark \ref{rem:truncate}. Throughout we use a model in which $Y$ depends only on a single variable (this is unknown to the algorithms). 
\eq{
Y_t \sim \begin{cases}
\bernoulli(\frac{1}{2}+\epsilon) & \text{if } X_1 = 1 \\
\bernoulli(\frac{1}{2}-\frac{q_1}{1-q_1}\epsilon) & \text{otherwise}\,.
\end{cases}
}

@@

\begin{figure}[h]
\centering
\includegraphics[width=.5\textwidth]{exp_regret_vs_m_N50_T400_s10000_20160205_2230}
\caption{Simple regret vs $m(\boldsymbol{q})$ for fixed horizon $T=400$ and number of variables $N = 50$}
\label{fig:simple_vs_m}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=.5\textwidth]{exp_regret_vs_T_N50_s10000_20160206_0533}
\caption{Simple regret vs horizon, $T$, with $N = 50$, $m=2$ and fixed $\epsilon = .3$}
\label{fig:simple_vs_T}
\end{figure}

### 2016-02-05 causal-bandit/icml-2015/general.tex
fixes

It is no longer optimal to divide our samples between observing to learn about multiple actions and explicit intervention to learn single ones. We need to incorporate the information we can learn about the reward for intervening on one variable from rounds in which we actually set a different variable.
Consider the graph in Figure \ref{fig:causalchain}, where each variable deterministically takes the value of its parent, $X_k = X_{k-1}$ 
for $k\in {2,\ldots,N}$ and $\P{X_1} = 0$. 
We can learn the reward for all the interventions $do(X_i = 1)$ simultaneously by selecting $do(X_1 = 1)$.

>>>>

We could naively generalize our approach for parallel bandits by observing for $T/2$ rounds, applying the truncated product factorization to write an expression for each $\P{Y|a}$ in terms of observational quantities and explicitly playing the actions $a$ for which the observational estimates were poor. However, the variance of the observational estimator for $a = do(X_i = j)$ can be high even if $\P{X_i = j}$ is large. In our example, suppose $\P{X_1=\frac{1}{2}}$ and $X_2 = X_1$ deterministically. $\P{X_2 = 1} = \frac{1}{2}$, however we will never observe $(X_2=1,X_1 = 0)$ so we cannot get a good estimate for $\P{Y|do(X_2=1)}$.

Additionally, it is no longer optimal to ignore the information we can learn about the reward for intervening on one variable from rounds in which we act on a different variable. Consider the graph in Figure \ref{fig:causalchain}, where each variable deterministically takes the value of its parent, $X_k = X_{k-1}$ 
for $k\in {2,\ldots,N}$ and $\P{X_1} = 0$. 
We can learn the reward for all the interventions $do(X_i = 1)$ simultaneously by selecting $do(X_1 = 1)$, (but not from $do()$).

%It is no longer optimal to divide our samples between observing to learn about multiple actions and explicit intervention to learn single ones. We need to incorporate the information we can learn about the reward for intervening on one variable from rounds in which we actually set a different variable.
%Consider the graph in Figure \ref{fig:causalchain}, where each variable deterministically takes the value of its parent, $X_k = X_{k-1}$ 
%for $k\in {2,\ldots,N}$ and $\P{X_1} = 0$. 
%We can learn the reward for all the interventions $do(X_i = 1)$ simultaneously by selecting $do(X_1 = 1)$.

### 2016-02-05 causal-bandit/icml-2015/simple-regret.tex
fixes

\begin{algorithm}[h]
\caption{Parallel Bandit Algorithm}\label{alg:simple}
\begin{algorithmic}[1]
\STATE {\bf Input:} Total rounds $T$ and $N$.
\FOR{$t \in 1,\ldots,T / 2$}
\STATE Perform empty intervention $do()$
\STATE Observe $\vec{X}_t$ and $Y_t$
\ENDFOR
\FOR{$a = do(X_i = x) \in \actions$}
\STATE Count times $X_i = x$ seen: $T_a = \sum_{t=1}^{T/2} \ind{X_{t,i} = x}$
\STATE Estimate reward: $\hat{\mu}_a = \frac{1}{T_a} \sum_{t=1}^{T/2} \ind{X_{t,i} = x} Y_t$ \\[0.2cm]
\STATE Estimate probabilities: $\hat{p}_a = \frac{2 T_a}{T}$,\,\, $\hat q_i = \hat p_{do(X_i = 1)}$
\ENDFOR
\STATE Compute $\hat{m} = m(\vec{\hat q})$ and $A = \set{a \in \actions \colon \hat{p}_a \leq \frac{1}{\hat m}}$.
\STATE Let $T_A := \frac{T}{2 |A|}$ be times to sample each $a\in A$.
\FOR{$a = do(X_i = x) \in A$}
\FOR{$t \in 1,\ldots,T_A$}
\STATE Intervene with $a$ and observe $Y_t$
\ENDFOR
\STATE Re-estimate $\hat{\mu}_a = \frac{1}{T_A} \sum_{t=1}^{T_A} Y_t$
\ENDFOR
\RETURN estimated optimal $\hat{a}^* \in \argmax_{a\in\actions} \hat{\mu}_a$
\end{algorithmic}
\end{algorithm}

@@

Note that algorithms designed for finite-armed bandits would explore interventions more uniformly and achieve a regret of $\Omega(\sqrt{N/T})$.
Since $m$ is typically much smaller than $N$ we anticipate that the new algorithm will significantly outperform classical bandit algorithms in
this setting.
We prove a lower bound on the simple regret that matches up to logarithmic factors the upper bound given in Theorem \ref{thm:uq-simple}

\begin{theorem}\label{thm:lower}
Let $\vec{q}$ be fixed and satisfies $m = m(\boldsymbol{q})$. Then for all strategies there exists a reward function such that
\eq{
\simpleregret 
%\geq \frac{\frac{m}{2e} - 1}{m} \sqrt{\frac{2m}{3T}} 
\in \Omega\left(\sqrt{\frac{m}{T}}\right)\,.
}
\end{theorem}

\ifsup
The proof of Theorems \ref{thm:uq-simple} and \ref{thm:lower} may be found in Sections \ref{sec:thm:uq-simple} and \ref{sec:thm:lower}.
\else
The proof of Theorems \ref{thm:uq-simple} and \ref{thm:lower} may be found in the supplementary material.
\fi

>>>>

\begin{algorithm}[h]
\caption{Parallel Bandit Algorithm}\label{alg:simple}
\begin{algorithmic}[1]
\STATE {\bf Input:} Total rounds $T$ and $N$.
\FOR{$t \in 1,\ldots,T / 2$}
\STATE Perform empty intervention $do()$
\STATE Observe $\vec{X}_t$ and $Y_t$
\ENDFOR
\FOR{$a = do(X_i = x) \in \actions$}
\STATE Count times $X_i = x$ seen: $T_a = \sum_{t=1}^{T/2} \ind{X_{t,i} = x}$
\STATE Estimate reward: $\hat{\mu}_a = \frac{1}{T_a} \sum_{t=1}^{T/2} \ind{X_{t,i} = x} Y_t$ \\[0.2cm]
\STATE Estimate probabilities: $\hat{p}_a = \frac{2 T_a}{T}$,\,\, $\hat q_i = \hat p_{do(X_i = 1)}$
\ENDFOR
\STATE Compute $\hat{m} = m(\vec{\hat q})$ and $A = \set{a \in \actions \colon \hat{p}_a \leq \frac{1}{\hat m}}$.
\STATE Let $T_A := \frac{T}{2 |A|}$ be times to sample each $a\in A$.
\FOR{$a = do(X_i = x) \in A$}
\FOR{$t \in 1,\ldots,T_A$}
\STATE Intervene with $a$ and observe $Y_t$
\ENDFOR
\STATE Re-estimate $\hat{\mu}_a = \frac{1}{T_A} \sum_{t=1}^{T_A} Y_t$
\ENDFOR
\RETURN estimated optimal $\hat{a}^*_T \in \argmax_{a\in\actions} \hat{\mu}_a$
\end{algorithmic}
\end{algorithm}

@@

Note that algorithms designed for finite-armed bandits would explore interventions more uniformly and achieve a regret of $\Omega(\sqrt{N/T})$.
Since $m$ is typically much smaller than $N$ we anticipate that the new algorithm will significantly outperform classical bandit algorithms in
this setting.
We prove a lower bound on the simple regret that matches up to logarithmic factors the upper bound given in Theorem \ref{thm:uq-simple}.

\begin{theorem}\label{thm:lower}
For all $T$, $\vec{q}$ and all strategies there exists a reward function such that
\eq{
\simpleregret 
\in \Omega\left(\sqrt{\frac{m(\vec{q})}{T}}\right)\,.
}
\end{theorem}

\ifsup
The proofs of Theorems \ref{thm:uq-simple} and \ref{thm:lower} may be found in Sections \ref{sec:thm:uq-simple} and \ref{sec:thm:lower} respectively.
\else
The proofs of Theorems \ref{thm:uq-simple} and \ref{thm:lower} follow by carefully analysing the concentration
of $\hat p_a$ and $\hat m$ about their true values and may be found in the supplementary material.
\fi
